---
title: "Spatiophylogenetic modelling: the end-Cretaceous extinction"
...

# Introduction

The aim of this part of the project is to use occurrence data of bivalves from across the Cretaceous–Palaeogene boundary to reconstruct extinction prevalence and potential drivers of this.

```{r renv_init, eval = FALSE}
renv::init()
```

```{r packages}
library(jsonlite)
library(ggthemes)
library(mapproj)
library(proj4)
library(rnaturalearth)
library(sf)
library(tidyverse)
library(viridis)
```

# Acquiring data

The data come from the [Paleobiology Database](https://paleobiodb.org/#/) (PBDB). Occurrences are downloaded though their [API](https://paleobiodb.org/data1.2/).

First we can download the data as a JSON file and save this to disc: this is our master data set. For this, I've gathered occurrences of Bivalvia from the Cretaceous and Paleogene with all of the occurrence, geographical, systematic, and ecological data.

```{r download_pbdb_data, eval = FALSE}
bivalve_pbdb_url <-
  paste0(
    "https://paleobiodb.org/data1.2/", # PBDB API URL
    "occs/", # occurrences
    "list.json?", # download a JSON file
    "base_name=Bivalvia&", # download bivalve occurrences
    "interval=Cretaceous,Paleogene&", # temporal range to include
    "show=full" # include full data
  )
bivalve_pbdb_file <- paste0(Sys.Date(), "-bivalve_pbdb_occurrences.json")
download.file(bivalve_pbdb_url, bivalve_pbdb_file, method = "curl")
```

The data file can then be loaded into R, which gives a list with length two: the first element (`$elapsed_time`) gives the original download time and the second element (`$records`) is a `data.frame` of the PBDB occurrence data.

```{r read_pbdb_data}
bivalve_pbdb_file <- paste0(Sys.Date(), "-bivalve_pbdb_occurrences.json")
bivalve_pbdb_data <- jsonlite::fromJSON(bivalve_pbdb_file)$records |>
  as_tibble()
```

We won't need data from all the columns, but the important few are:

* `oid`: occurrence ID, a unique reference number.
* `idn`: the originally identified name.
* `tna`: the currently accepted name.
* `rnk`: taxonomic rank of the accepted name.
* `oei`, `oli`: earliest and latest occurrence interval names.
  - `eag` and `lag` are ages at the beginning and end of these intervals, taken from the [*Geologic Time Scale*](https://timescalecreator.org/index/index.php), but may want to update to a recent version.
* `fml`: family of the occurring taxon.
  - higher taxon levels are also included.
* `lng`, `lat`: modern longitude and latitude of the occurrence.
  - `pln` and `pln` are the reconstructed palaeocoordinates, but it may be worth re-doing these to confirm or with a different tectonic model.
* `env`: the palaeoenvironment, useful to check localities and habitats, exclude freshwater species etc. 
* Various [palaeocological/ecospace data](https://paleobiodb.org/data1.2/general/ecotaph_doc.html#Ecospace), such as habitat depth (`jlh`), general environment (`jev`), feeding mode (`jdt`).
* `jco`: composition/mineralogy of shells.

A full list of the fields is at the bottom of [this page](https://paleobiodb.org/data1.2/occs/list_doc.html#response).

# Querying data #

It's a good check to have a quick look at the data to check its size and what different values are contained within. Most of the PBDB data are categorical or strings, which makes it a little more difficult (you can't plot a bar chart for instance, except you can). A few things to check:

* How many records are identified to species level? genus? family?
* How many different species/genera/families are include?
* How many records don't have accepted values?


```{r check_species}
bivalve_pbdb_data$records |>
  filter(rnk == 3) |>
  group_by(tna) |>
  summarize(n())
```

Or a more fun thing to plot the occurrences from different families.

```{r plot_family_occurrences}
family_occurrences <-
  bivalve_pbdb_data$records |>
    filter(rnk == c(3, 4, 5)) |>
    group_by(fml) |>
    summarize(family_occurrences = n())
family_occurrences |>
  ggplot(aes(x = reorder(fml, desc(family_occurrences)), y = family_occurrences)) +
    geom_col()
family_occurrences |>
  slice_max(family_occurrences, n = 10)
```

## Data quality ##

Now we can do a more thorough look through the data to check that it's correct, or at least in the form we expect it to be. First is the relatively easy one: checking for numbers in columns that should have the, such as rnk, lng, and lat. On import this column is started as a character string in the tibble. We can check whether any of the records are not proper numbers by searching for characters not found in a number: i.e. not digits, decimal points, or minus signs.

```{r data_cleaning}
# search for strings that don't contain a digit, full stop/decimal point, or hyphen/minus, i.e. those that aren't a number
number_as_character_regexp <- "[^[[:digit:]]\\.\\-]"

check_numeric_column <- function(column, data) {
  column_vector <- pull(data, column)
  string_match <- str_which(column_vector, number_as_character_regexp)
  if (length(string_match) == 0) {
    print(paste(column, ": All items are (probably) numbers."))
  } else {
    print(paste(column, ": Some items (probably) aren't numbers."))
    string_match
  }
}

c("lng", "lat", "rnk") |>
  walk(check_numeric_column, data = bivalve_pbdb_data)
```

Assuming this above comes out tickety-boo, then we can simply replace and reformat those columns to numeric values. It's always a good plan to create a new object doing this so that you don't lose the original data, just in case things go wrong later.

```{r reformatting_numeric_data}
bivalve_cleaned <-
  bivalve_pbdb_data |>
  mutate(across(c(rnk, lng, lat), as.numeric))
```

The same should be done for the columns we expect to have character strings in, i.e. the accepted name and family column.


# Mapping data #

We can also do more fun things like plot the occurrences on a map.

```{r map_occurrences}
world_map <-
  ne_countries(returnclass = "sf") |>
  st_transform(crs = "+proj=moll")

occurrences_to_plot <-
  bivalve_cleaned |>
  filter(rnk == 3) |>
  slice_sample(n = 1000)

projected_coords <- 
project(occurrences_to_plot[, c("lng", "lat")], proj = "+proj=moll")
occurrences_to_plot <-
  occurrences_to_plot |>
  mutate(
    lng = projected_coords$x,
    lat = projected_coords$y
  )

  ggplot() +
    geom_sf(data = world_proj, inherit.aes = FALSE) +
    geom_point(data = occurrences_to_plot, mapping = aes(x = lng, y = lat, colour = fml)) +
    scale_colour_viridis(discrete = TRUE) +
    coord_sf(crs = st_crs(world_proj)) +
    theme_map() +
    theme(
      legend.position = "none",
      plot.background = element_blank(),
    )
```
